---
title: "Feature Engineering"
format: html
---

# Introduction  
The practice of finding better representations of predictors (i.e., features) to improve predictive performance is referred to as “feature engineering.”

In other words, sometimes the predictor variables at hand are not in themselves the best representation to explain a predicted variable.  

In these cases, we can use these original predictor variables to derive new, more informative secondary variables.  


# Learning objectives  
Our learning objectives are to:  

  - Apply feature engineering concepts to Daymet daily weather data  
  - Export the feature-engineered data set for further use 

# Setup  
```{r}
#| message: false
#| warning: false

library(tidyverse)
```

```{r}
fieldweather <- read_csv("../data/fieldweatherdata.csv")

fieldweather
```

Our data set contains **254,770** rows comprised by **698** site-years and **365** days/site-year worth of weather data.  

# EDA  
```{r}
summary(fieldweather)
```

Let's create some density plots to explore the weather data distributions.  
```{r}
fieldweather %>%
  pivot_longer(cols = dayl_s:vp_pa) %>%
  ggplot(aes(x = value)) + 
  geom_density() +
  facet_wrap(~name, scales = "free") 

```

# Feature engineering  
Let's take another look at the data at hand:
```{r}
fieldweather 
```

Currently, we have weather data for each site-year at a **daily** time interval.  

Although this level of temporal resolution is great, plant outputs like yield or quality rarely respond to weather events that occur in one given day (unless it is a severe event like flooding or tornado).  

Rather, plants respond to the cumulative effects of weather across many days. 

How could we use this understanding to create new features from the current ones?    

## Quiz - go to eLC  










There are **two components** that we can consider when feature engineering weather data:  

1. Summarizing time **window size**  
  - Weekly  
  - Monthly  
  - Every 2 months  
  - Growing season  
  - Based on crop growth stages  
  - Other?  
  
2. Summarizing **function**  
  - Mean  
  - Median  
  - Minimum  
  - Maximum  
  - Standard deviation  

In the next section, let's use the following:  
  - Summarizing time window size: **monthly**  
  - Summarizing function: **mean** or **sum**  

## Quiz - go to eLC  

Let's take a look into date formats in R:  
https://www.r-bloggers.com/2013/08/date-formats-in-r/ 

A great package to work with dates and times in R is called **lubridate**, see vignette here: https://lubridate.tidyverse.org/index.html  

Next, let's create a column containing the month information.  

```{r fe_month}
fe_month <- fieldweather %>%
  # Selecting needed variables
  dplyr::select(year, site, lat, lon,
                strength_gtex,
                yday,
                dayl.s = dayl_s, 
                prcp.mm = prcp_mm_day,
                srad.wm2 = srad_w_m_2, 
                tmax.c = tmax_deg_c, 
                tmin.c = tmin_deg_c,
                vp.pa = vp_pa
                ) %>%
  # Creating a date class variable  
  mutate(date_chr = paste0(year,"/",yday)) %>%
  mutate(date = as.Date(date_chr,"%Y/%j")) %>%
  # Extracting month from date  
  mutate(month = month(date)) %>%
  mutate(month_abb = month(date, label = T))

fe_month
```

Now, let's summarize daily weather variables based on month.  
```{r fe_month_sum}
fe_month_sum <- fe_month %>%
  #because every colunm grouped will show, so we need include strength,which is the y
  group_by(year, site, month_abb, strength_gtex) %>%
  summarise(across(.cols = c(dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa
                             ),
                   .fns = mean,
                   .names = "mean_{.col}"
                   ),
            across(.cols = prcp.mm,
                   .fns = sum,
                   .names = "sum_{.col}")
            ) %>%
  ungroup()


fe_month_sum
```

8,376 rows because 698 site-years x 12 months.  

Let's check tmax.c and prcp.mm for the first site-year and month.  

```{r}
fe_month %>%
  filter(year == 1980 & 
           site == "Altus, OK" &
           month_abb == "Jan") %>%
  summarise(tmax.c = mean(tmax.c),
            prcp.mm = sum(prcp.mm))
```

Now, what if we wanted to have month as part of the column name instead?  

```{r fe_month_sum_wide}
fe_month_sum_wide <- fe_month_sum %>%
  pivot_longer(mean_dayl.s:sum_prcp.mm) %>%
  mutate(varname = paste0(name, "_",month_abb)) %>%
  dplyr::select(-name,-month_abb) %>%
  pivot_wider(names_from = varname,
              values_from = value) %>%
  # Rounding to one decimal point
  mutate(across(c(3:75), ~round(.,1)))

fe_month_sum_wide  
```
Notice how we are back at **698** rows.  

For each site-year, we have one piece of weather information for each of the weather variables (as opposed to the original Daymet data that had 365 rows per site-year).  

Let's explore our newly engineered variables.  

# EDA round 2  
Let's make a ridge plot to visualize the distribution of one variable over months.  

```{r}
#install.packages("ggridges")
library(ggridges)

ggplot(data = fe_month_sum,
       aes(x = mean_tmax.c,
           y = month_abb,
           fill = stat(x))
       ) + geom_density_ridges_gradient(scale = 2,
                                        rel_min_height = 0.01)  +
  scale_fill_viridis_c(option = "C",alpha = .8) +
  theme(legend.position = "none") +
  labs(x = "Monthly-averaged maximum temperature (C)")
```

Now let's do that for all variables.  
```{r}
finalplots <- fe_month_sum %>%
  pivot_longer(mean_dayl.s:sum_prcp.mm) %>%
  group_by(name) %>%
  nest() %>%
  mutate(plot = map2(data,name,
                     ~ggplot(data = .x,
       aes(x = value ,
           y = month_abb,
           fill = stat(x))
       ) + geom_density_ridges_gradient(scale = 2,
                                        rel_min_height = 0.01)  +
  scale_fill_viridis_c(option = "C",alpha = .8) +
  theme(legend.position = "none") +
  labs(x = .y,
       y = "Month"
       )
                       
                       ))
  
finalplots$plot
```

```{r}
#| message: false

finalplots$plot
```

# Exporting  
```{r}
write_csv(fe_month_sum_wide,
          "../data/weather_monthsum.csv")
```

# Summary  
In this exercise, we:  
  - Imported the original **7** weather variables from Daymet  
  - Feature engineered a total of **72** secondary variables by applying a summarizing window size of **month** and a summarizing function of **mean or sum**.  
  - Explored the data distribution of the new variables.  
  - Exported to file to be used in subsequent exercises.




  
