---
title: "Data Wrangling"
format: html
editor: Ming/Ishwari
---

# 1) Setup

Prepare the packages and raw materials that we need for data wrangling.

```{r Setup, message=FALSE}
# Loading packages (every time you open R)
#install.packages("nasapower") #
#install.packages("beepr")
library(readxl) # to read excel files
library(janitor) # to clean data
library(dplyr) # wranglingdd
library(tidyr) # wrangling
library(readr) # to export csv
library(nasapower) # for NASA POWER API
library(purrr) # for mapping
library(beepr) # for beep
library(stringr)
library(USAboundaries) # for US state boundaries
library(sf) # for US map
library(lubridate) # for date math
.libPaths()
```

```{r Read files}
getwd() #get your working directory
Meta <- read_csv("../data/training/training_meta.csv")
Soil <- read_csv("../data/training/training_soil.csv")
Trait <- read_csv("../data/training/training_trait.csv")
Meta <- read_csv("../data/training/training_meta.csv")
glimpse(Meta)
glimpse(Soil)
glimpse(Trait)

```

# 2) Open source data

```{r test api }
# get weather data from NASA Power based on year, site, lon, lat in meta data

meta_unique <- Meta %>%
  distinct(site, year, latitude, longitude, previous_crop)
meta_cleaned <- clean_names(Meta)


site_coord <- meta_cleaned %>%
  group_by(site) %>%
  summarise(
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    .groups = "drop"
  )
glimpse(meta_unique) 
# Example of R api 
#https://docs.ropensci.org/nasapower/articles/nasapower.htm

# learn api
help(package = "nasapower")



# test api
weather_test <- get_power(
  community = "AG",   # agriculture
  lonlat = c(-83.4, 35.2),  # 经度, 纬度
  dates = c("2014-01-01", "2023-12-31"),
  pars = c(
  "T2M",
  "T2M_MAX",
  "T2M_MIN",
  "PRECTOTCORR",
  "ALLSKY_SFC_SW_DWN"
)  # mean tmp, max, min, precipitation, radiation
)

glimpse(weather) 
```

```{r create a function}
get_weather <- function(lon, lat, yr, site_name ) {
  cat("Running:", site_name, yr, "\n")
  x <- get_power(
    community = "AG",
    lonlat = c(lon, lat),
    dates = c(paste0(yr, "-01-01"), paste0(yr, "-12-31")),
    pars = c(
      "T2M",
      "T2M_MAX",
      "T2M_MIN",
      "PRECTOTCORR",
      "ALLSKY_SFC_SW_DWN"
    ),
    temporal_api = "daily"
  )
  
  Sys.sleep(1)  # avoid be banned by api limit 
   return(x) # return weather data frame
}


```

```{r test func}
weather_test_func <- meta_unique %>% slice(1:5) %>%
  mutate(
    weather = pmap(
      list(longitude, latitude, year, site),
      ~ get_weather(..1, ..2, ..3, ..4)
    )
  ) %>%
  tidyr::unnest(weather)

```

```{r iteration eval=FALSE}
#After downloading it, eval=FALSE this chunk.
weather_all_year <- meta_unique %>%
  mutate(
    weather = pmap(
      list(longitude, latitude, year, site),
      ~ get_weather(..1, ..2, ..3, ..4)
    )
  ) %>%
  tidyr::unnest(weather)
 #beep
  beep()
```

```{r check the number of site for each year}
meta_unique %>%
  count(year)
```

```{r save as csv}
write_csv(weather_all_year, "../data/weather_raw.csv")
```



# 3) Data wrangling

## a. On training data

```{r clean_names }
meta_cleaned <- clean_names(Meta)
soil_cleaned <- clean_names(Soil)
trait_cleaned <- clean_names(Trait)

glimpse(soil_cleaned)

```

Trying to merge three csv. But lots of problems on site-year. So we have to deal with it before merge. Different code plus num(e.g TXH1, TXH2), means they are different site, means different lan and lon.

later suffix either lower case a/b or late/early(e.g "MOH1a", "MOH1-rep2") means different rep/block info, but doesn't affect the lan and lon info.

```{r handling site-year}
# clean soil, separate site-year
soil_separate <- soil_cleaned %>%
  # "_(?=[^_]+$)" means look for the last "_" from right side to do the separation.
  separate(site, into = c("site", "year"), sep = "_(?=[^_]+$)") %>%
  mutate(year = as.integer(year))


# trait,soil, and meta all have different site-year. check the number
trait_key <- trait_cleaned %>%
  distinct(site, year)
# 270 rows

meta_key <- meta_cleaned %>%
  distinct(site, year)
# 251 rows

soil_key <- soil_separate %>%
  distinct(site, year)
# 186 keys


# avoid many-to-many merge
## check meta lat,lon base on site-year info
meta_cleaned %>%
  count(site, year) %>%
  filter(n > 1)

meta_check_site_year<-meta_cleaned %>%
  group_by(site, year) %>%
  summarise(n = n(),
            n_distinct_lat = n_distinct(latitude),
            n_distinct_lon = n_distinct(longitude))
## lan and lon all same, so we can use distinct as long as they're are from same site code, even if the lan and lon differ slightly. But we have to choose one set of lan and lon from the same site code, to avoid the many-to-many merge problem.

# average the lon and lat for the same site code
meta_unique <- meta_cleaned %>%
  group_by(site, year) %>%
  summarise(
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    previous_crop = first(previous_crop),
    .groups = "drop"
  )


# do the same to soil data
soil_unique <- soil_separate %>%
  group_by(site, year) %>%
  summarise(
    soilp_h = mean(soilp_h, na.rm = TRUE),
    om_pct = mean(om_pct, na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups = "drop"
  )

# now we have the soil and meta data with distinct site year so that we can avoid that many-to-many problem when merge to trait data 

# set same site for 3 csv

# test regular expression
sub("^([A-Za-z]+[0-9]+).*", "\\1",
    c("TXH1-Late", "MOH1a", "MOH1-rep2", "IAH1"))
#  "TXH1" "MOH1" "MOH1" "IAH1"

trait2 <- trait_cleaned %>%
  rename(detailed_site = site) %>%
  mutate(
    site = detailed_site %>%
    sub("^([A-Za-z]+[0-9]+).*", "\\1", x = .) # remove a/b/c,rep, later/early
      
  ) %>% 
  # move to 3rd col
  relocate(site, .after = detailed_site)

# check trait2, row #943, IAH1a -> IAH1, it worked.
# row # 49901 2017 TXH1-Early -> TXH1, it worked.

```

Now, we make the site clearly separate from detailed site in trait data. And we make site unique in meta and soil data. We can start merge.

```{r  merge data & check }

# merge
data_merged <- trait2 %>%
  left_join(meta_unique, by = c("site" = "site", "year" = "year")) %>%
  left_join(soil_unique, by = c("site" = "site", "year" = "year"))

# check many-to-many problem
nrow(trait2) == nrow(data_merged)
# True means we did it right, there's no many-to-many problem happened.


# overwrite with rep info in soil data
soil_rep <- soil_separate %>%
  mutate(
    site_base = str_remove(site, "_\\d+$"),
    rep_soil = as.integer(str_extract(site, "(?<=_)\\d+$"))
  ) %>%
  filter(!is.na(rep_soil)) %>%
  select(site_base, year, rep_soil, soilp_h, om_pct, soilk_ppm, soilp_ppm)

data_fix <- data_merged %>%
  left_join(
    soil_rep,
    by = c("site" = "site_base", "year" = "year", "replicate" = "rep_soil"),
    suffix = c("", "_rep")
  )

data_fix <- data_fix %>%
  mutate(
    soilp_h = coalesce(soilp_h_rep, soilp_h),
    om_pct = coalesce(om_pct_rep, om_pct),
    soilk_ppm = coalesce(soilk_ppm_rep, soilk_ppm),
    soilp_ppm = coalesce(soilp_ppm_rep, soilp_ppm)
  ) %>%
  select(-ends_with("_rep"))
# read the row#63403 &63404, the soil is different now. Means we give the different rep to trait data.

# save as merged data
write_csv(data_fix, "../data/training/training_merged.csv")
```



This means we merge correctly, even though we lack some site-year info initially from meta and soil data.

Then we can do some data wrangling & feature engineering on weather data. 

## b. On testing data 

We should do the same to testing data as to training data.

```{r clean_names }
test_soil_cleaned <- read_csv("../data/testing/testing_soil.csv") %>%
  clean_names()
test_meta_cleaned <- read_csv("../data/testing/testing_meta.csv") %>%
  clean_names()
test_submission_cleaned <- read_csv("../data/testing/testing_submission.csv") %>%
  clean_names()



glimpse(test_soil_cleaned)

```

```{r separate year and site for soil and meta}
test_soil_sep <- test_soil_cleaned %>%
  # "_(?=[^_]+$)" means look for the last "_" from right side to do the separation.
  separate(site, into = c("site", "year"), sep = "_(?=[^_]+$)") %>%
  mutate(year = as.integer(year)) 

# get unique site year, by average
test_soil_sep2 <- test_soil_sep %>%
  group_by(site, year) %>%
  summarise(
    soilp_h = mean(soilp_h, na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    om_pct = mean(om_pct, na.rm = TRUE),
    .groups = "drop"
  )
# now it's 16 rows


# test_meta all rows unique, so we don't need this step for test_meta


```

```{r merge & check}
# merge
test_data_merged <- test_submission_cleaned %>%
  left_join(test_soil_sep2, by = c("site" = "site", "year" = "year")) %>%
  left_join(test_meta_cleaned, by = c("site" = "site", "year" = "year"))
# check many-to-many problem
nrow(test_data_merged) == nrow(test_submission_cleaned)
# True means we did it right, there's no many-to-many problem happened.

 
# save as merged data
write_csv(test_data_merged, "../data/testing/testing_merged.csv")
```

# 4) Feature engineering on weather data

## a. Download weather data again

Based on updated view on the merged trait data, we have to download weather again for all the unique site-year across whole 2013-2024 12 years.

```{r read the data,check the site-year match }
#check the distint site-year of weather
weather <- read.csv("../data/weather_raw.csv")
weather_unique_siteyear <- weather %>%
  distinct(site, year) 
# 251 rows

#check the distint site-year of merged data
data_merged <- read.csv("../data/training/training_merged.csv")
train_data_unique_siteyear<-  data_merged %>%
  distinct(site, year)
# 264 rows

# then we need to download again weather data about all the unique site 
train_data_unique_site <- data_merged %>%
  distinct(site, latitude,longitude)
# 39 rows,
# we need all these site across 2014-2023 years 
site_coord <- data_merged %>%
  group_by(site) %>%
  summarise(
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    .groups = "drop"
  ) %>%  
  # remove na lat and lon
  filter(!is.na(latitude), !is.na(longitude))


```

```{r create a function}
get_weather <- function(lon, lat, site_name) {
  cat("Running:", site_name, "\n")
  
  x <- get_power(
    community = "AG",
    lonlat = c(lon, lat),
    dates = c("2013-01-01", "2024-12-31"),
    pars = c(
      "T2M",
      "T2M_MAX",
      "T2M_MIN",
      "PRECTOTCORR",
      "ALLSKY_SFC_SW_DWN"
    ),
    temporal_api = "daily"
  )
  
  x$site <- site_name
  
  Sys.sleep(0.5)
  
  return(x)
}


```

```{r test func}
weather_test_func <- site_coord %>%
  slice(1:5) %>%
  mutate(
    weather = pmap(
      list(longitude, latitude, site),
      ~ get_weather(..1, ..2, ..3)
    )
  ) %>%
  tidyr::unnest(weather)
  #beep
  beep()
```

```{r iteration eval=FALSE}
#After downloading it, eval=FALSE this chunck
weather_raw3<- pmap_dfr(
  list(site_coord$longitude, site_coord$latitude, site_coord$site),
  get_weather
)
  #beep
  beep()
  
  # save data 
  write_csv(weather_raw3, "../data/weather_allsites_12years3.csv")
```



Until now, we get the new weather data, let's start feature engineering.  

## b. EDA for train data
```{r EDA for the training merged data }
train_data <- read_csv("../data/training/training_merged.csv")
weather <- read_csv("../data/weather_allsites_12years.csv")
# get cord for unique site
field <- train_data %>%
  group_by(site) %>%
  summarise(
    lat = mean(latitude, na.rm = TRUE),
    lon = mean(longitude, na.rm = TRUE),
    yield_mean = mean(yield_mg_ha, na.rm = TRUE)
  )
 
summary(train_data)
summary(weather)
```

How many rows for weather? 38 unique sites × 4383 days(3 leap years) = 166,554 rows

What is the statistical distribution of fiber strength?

```{r yield statistical distribution}
ggplot(data = train_data,
       aes(x = yield_mg_ha)
       ) + 
  geom_density()
```

Train data is nearly normal distributed, a slightly right skewed.

What's site location?

```{r  show on map }
states <- us_states() %>%
  filter(!(state_abbr %in% c("PR","AK","HI")))

states

ggplot() +
  geom_sf(data = states) + 
  geom_point(data = field,
    aes(x = lon,
        y = lat,
        color = yield_mean
)
)
# it's pretty interesting that we have a site in Europe. After checking the meta data, we find that there're more site-year and slightly different lat and lon on this same site location, which means it's not a mistake. It is designed. And according to the way we collect the weather data, it's valid data cuz NASA provide the global data.

```

## c. EDA for weather
Let's create some density plots to explore the weather data distributions.

```{r EDA for weather data}

weather2 <- weather %>%
  pivot_longer(cols = T2M:ALLSKY_SFC_SW_DWN) %>%
  ggplot(aes(x = value)) + 
  geom_density() +
  facet_wrap(~name, scales = "free") 
weather2
```

On weather data, we have radiation, precipitation, temp mean, min and max. And we have the plant and harvest date on training data. There are **two components** that we can consider when feature engineering weather data:\
1. Summarizing time **window size**\
- Weekly ✅ - Monthly\
- Every 2 months\
- Growing season ✅ - Based on crop growth stages ✅

2.  Summarizing **function**\

-   Mean ✅
-   Sum ✅
-   Standard deviation

Before we decide, let's check the plant and harvest date.

```{r check plant and harvest date}
train_data <- read_csv("../data/training/training_merged.csv")

train_data2 <- train_data %>%
  mutate(
    date_planted = as.Date(date_planted, format = "%m/%d/%y"),
    date_harvested = as.Date(date_harvested, format = "%m/%d/%y"),
    season_length = as.numeric(date_harvested - date_planted)
  )

train_data2

```

Notice every site-year plant/harvest at different date, and the period of growth also different. So we decide to use relative growth period as the time scale for feature engineering on weather data. So we can set **10 quantile stages** for corn.

Get these values for every stage: **GDD** , Growing Degree Days: GDD = ((Tmax + Tmin) / 2) - Tbase For corn, let's set: Tbase = 10°C Tmax cap = 30°C Tmin floor = 10°C

**sum of radiation**

**sum of precipitation**

**mean of T2M/max/min**

```{r  phenology info}
# phenology data frame
pheno_date <- train_data2 %>%
  group_by(site, year) %>%
  summarise(
    date_planted = date_planted,
    date_harvested = date_harvested ,
    .groups = "drop"
  )

 pheno_date %>%
  count(site, year, date_planted, date_harvested) %>%
  arrange(site, year)
# check the result to find out that in same site-year, harvest date might slightly changed 1-3days. This change is from rep and block difference, since this info is not provided in test data, so we decide to ignore it.
 
# we choose the max dates for same site year, reason is harvest may last for couples of days.  
 pheno_date2 <- pheno_date %>%
  group_by(site, year) %>%
  summarise(
    date_planted = first(date_planted),
    date_harvested = max(date_harvested, na.rm = TRUE),
    .groups = "drop"
  )
# 264 rows

  

```

```{r filter out the doy of growth period in weather data}
weather3 <- weather %>% 
  clean_names() %>%
  left_join(pheno_date2, by = c("site", "year")) %>%
  filter(yyyymmdd >= date_planted & yyyymmdd <= date_harvested)
# weather3: one row for one day, and start from plant date finish at harvest date for each site-year.
weather3 %>%
  distinct(site, year) %>%
  nrow()
# 262 rows, means 2 rows of site-year missing.
anti_join(
  pheno_date2 %>% distinct(site, year),
  weather3 %>% distinct(site, year),
  by = c("site", "year")
)
# These two site-year do not have correct lat and lon info: NYH1 in 2015, TXH4 in 2019.
```

```{r divide into 10 staged based on quantile}
# calculate the stage progress
weather3 <- weather3 %>%
  mutate(
    season_length = as.numeric(date_harvested - date_planted),
    dap = as.numeric(yyyymmdd - date_planted),
    # dap:days after planting
    progress = dap / season_length
  )

# create 10 stages
weather3 <- weather3 %>%
  mutate(
    stage = cut(
      progress,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE,
      labels = paste0("s", 1:10)
    )
  )



```

```{r save weather stage}
write_csv(weather3,
          "../data/weather_stage.csv")
```


Finally we can calculate the GDD now based on the growing stage. **GDD** , Growing Degree Days: GDD = ((Tmax + Tmin) / 2) - Tbase

```{r calculate the GDD}
# the reason of this is, corn is not growing under 10 degree, only growing from 10-30, not growing over 30 degree.
weather4 <- weather3 %>%
  mutate(
    tmax_adj = pmin(t2m_max, 30),
    tmin_adj = pmax(t2m_min, 10),
    gdd = pmax(((tmax_adj + tmin_adj) / 2) - 10, 0)
  )

```

```{r get weather feature based on stage}
# group by stage
weather_stage <- weather4 %>%
  group_by(site, year, stage) %>%
  summarise(
    gdd = sum(gdd, na.rm = TRUE),
    prcp = sum(prectotcorr, na.rm = TRUE),
    rad  = sum(allsky_sfc_sw_dwn, na.rm = TRUE),
    t2m  = mean(t2m, na.rm = TRUE),
    t2max = max(t2m_max, na.rm =TRUE),
    t2min = min(t2m_min, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r weathre feature wide}
weather_stage_wide <- weather_stage %>%
  pivot_wider(
    names_from = stage,
    values_from = c(gdd, prcp, rad, t2m, t2max, t2min),
    values_fill = 0
  )
# still the 262 rows, means the number of unique site-year, same rows
```



## d. Feature refinement
Now we have stage wide weather data, but we could do more like get the sum of gdd across whole growth period, which means the sum of 10 stages.
```{r feature refinement }
weather_stage_wide2 <- weather_stage_wide %>%
  rowwise() %>%
  mutate(
    # add total gdd, percipitation, radiation across 10 stages.
    total_gdd  = sum(c_across(gdd_s1:gdd_s10), na.rm = TRUE),
    total_prcp = sum(c_across(prcp_s1:prcp_s10), na.rm = TRUE),
    total_rad  = sum(c_across(rad_s1:rad_s10), na.rm = TRUE),
    # add tem mean, max, min across 10 stages.
    avg_t2m    = mean(c_across(t2m_s1:t2m_s10), na.rm = TRUE),

    max_t2max  = max(c_across(t2max_s1:t2max_s10), na.rm = TRUE),
    min_t2min  = min(c_across(t2min_s1:t2min_s10), na.rm = TRUE),
    
    # do more stages sparation like 3 stages: early, mid and late.
    # here is gdd
    gdd_early  = sum(c_across(gdd_s1:gdd_s3), na.rm = TRUE),
    gdd_mid    = sum(c_across(gdd_s4:gdd_s7), na.rm = TRUE),
    gdd_late   = sum(c_across(gdd_s8:gdd_s10), na.rm = TRUE),
    # do prcp also
    prcp_early = sum(c_across(prcp_s1:prcp_s3), na.rm = TRUE),
    prcp_mid   = sum(c_across(prcp_s4:prcp_s7), na.rm = TRUE),
    prcp_late  = sum(c_across(prcp_s8:prcp_s10), na.rm = TRUE)
  ) %>%
  ungroup()

```

```{r  more weather feature plans}
# We ignore the sum of radiation of 3 stages last chunk. Because it might now as important as the prcp and gdd. We can put this into another file, and see the result difference when training model.
weather_stage_wide3 <- weather_stage_wide %>%
  rowwise() %>%
  mutate(
    # add total gdd, percipitation, radiation across 10 stages.
    total_gdd  = sum(c_across(gdd_s1:gdd_s10), na.rm = TRUE),
    total_prcp = sum(c_across(prcp_s1:prcp_s10), na.rm = TRUE),
    total_rad  = sum(c_across(rad_s1:rad_s10), na.rm = TRUE),
    # add tem mean, max, min across 10 stages.
    avg_t2m    = mean(c_across(t2m_s1:t2m_s10), na.rm = TRUE),

    max_t2max  = max(c_across(t2max_s1:t2max_s10), na.rm = TRUE),
    min_t2min  = min(c_across(t2min_s1:t2min_s10), na.rm = TRUE),
    
    # do more stages sparation like 3 stages: early, mid and late.
    # here is gdd
    gdd_early  = sum(c_across(gdd_s1:gdd_s3), na.rm = TRUE),
    gdd_mid    = sum(c_across(gdd_s4:gdd_s7), na.rm = TRUE),
    gdd_late   = sum(c_across(gdd_s8:gdd_s10), na.rm = TRUE),
    # do prcp also
    prcp_early = sum(c_across(prcp_s1:prcp_s3), na.rm = TRUE),
    prcp_mid   = sum(c_across(prcp_s4:prcp_s7), na.rm = TRUE),
    prcp_late  = sum(c_across(prcp_s8:prcp_s10), na.rm = TRUE),
    # do rad also
    rad_early = sum(c_across(rad_s1:rad_s3), na.rm = TRUE),
    rad_mid   = sum(c_across(rad_s4:rad_s7), na.rm = TRUE),
    rad_late  = sum(c_across(rad_s8:rad_s10), na.rm = TRUE)
  ) %>%
  ungroup()

```


```{r Exporting weather wide}
write_csv(weather_stage_wide2,
          "../data/weather_stage_wide.csv")
```

```{r Exporting weather wide-plan2}
write_csv(weather_stage_wide3,
          "../data/weather_stage_wide_plan2.csv")
```

Now we have merged train data, merged test data, and weather wide data.

## e. Repeat for weather test data
But we find problems when move weather to test data. Because it's the weather for train data.

We should have done this before, but since we just find out this problem on this stage, We'll put it here. It just repeats the process from 4)-c to 4)-d. 


Stage dividing to test date. Basically same thing to test data.

```{r  phenology info for test set}
# phenology data frame
pheno_date_test <- test_stage_base %>%
  group_by(site, year) %>%
  summarise(
    date_planted = date_planted,
    date_harvested = date_harvested ,
    .groups = "drop"
  )
 pheno_date_test %>%
  count(site, year, date_planted, date_harvested) %>%
  arrange(site, year)
# check the result to find out that in same site-year, harvest date might slightly changed 1-3days. This change is from rep and block difference, since this info is not provided in test data, so we decide to ignore it.
 
# we choose the max dates for same site year, reason is harvest may last for couples of days.  
 pheno_date_test2 <- pheno_date_test %>%
  group_by(site, year) %>%
  summarise(
    date_planted = first(date_planted),
    date_harvested = max(date_harvested, na.rm = TRUE),
    .groups = "drop"
  )
# 23 rows

```

```{r filter out the doy of growth period in weather data}
weather_test2 <- weather_test %>% 
  clean_names() %>%
  left_join(pheno_date_test2, by = c("site", "year")) %>%
  filter(yyyymmdd >= date_planted & yyyymmdd <= date_harvested)
# weather3: one row for one day, and start from plant date finish at harvest date for each site-year.
weather_test2 %>%
  distinct(site, year) %>%
  nrow()
# 23 rows
anti_join(
  pheno_date_test %>% distinct(site, year),
  weather_test2 %>% distinct(site, year),
  by = c("site", "year")
)
# 0
```

```{r divide into 10 staged based on quantile}
# calculate the stage progress
weather_test3 <- weather_test2 %>%
  mutate(
    season_length = as.numeric(date_harvested - date_planted),
    dap = as.numeric(yyyymmdd - date_planted),
    # dap:days after planting
    progress = dap / season_length
  )

# create 10 stages
weather_test4 <- weather_test3 %>%
  mutate(
    stage = cut(
      progress,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE,
      labels = paste0("s", 1:10)
    )
  )



```



```{r calculate the GDD}
# the reason of this is, corn is not growing under 10 degree, only growing from 10-30, not growing over 30 degree.
weather_test5 <- weather_test4 %>%
  mutate(
    tmax_adj = pmin(t2m_max, 30),
    tmin_adj = pmax(t2m_min, 10),
    gdd = pmax(((tmax_adj + tmin_adj) / 2) - 10, 0)
  )

```

```{r get weather feature based on stage}
# group by stage
weather_test_stage <- weather_test5 %>%
  group_by(site, year, stage) %>%
  summarise(
    gdd = sum(gdd, na.rm = TRUE),
    prcp = sum(prectotcorr, na.rm = TRUE),
    rad  = sum(allsky_sfc_sw_dwn, na.rm = TRUE),
    t2m  = mean(t2m, na.rm = TRUE),
    t2max = max(t2m_max, na.rm =TRUE),
    t2min = min(t2m_min, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r weathre feature wide}
weather_test_stage_wide <- weather_test_stage %>%
  pivot_wider(
    names_from = stage,
    values_from = c(gdd, prcp, rad, t2m, t2max, t2min),
    values_fill = 0
  )
# still the 23 rows, means the number of unique site-year, same rows
```

```{r feature refinement for weather_test_stage_wide}
weather_test_stage_wide2 <- weather_test_stage_wide %>%
  rowwise() %>%
  mutate(
    # add total gdd, percipitation, radiation across 10 stages.
    total_gdd  = sum(c_across(gdd_s1:gdd_s10), na.rm = TRUE),
    total_prcp = sum(c_across(prcp_s1:prcp_s10), na.rm = TRUE),
    total_rad  = sum(c_across(rad_s1:rad_s10), na.rm = TRUE),
    # add tem mean, max, min across 10 stages.
    avg_t2m    = mean(c_across(t2m_s1:t2m_s10), na.rm = TRUE),

    max_t2max  = max(c_across(t2max_s1:t2max_s10), na.rm = TRUE),
    min_t2min  = min(c_across(t2min_s1:t2min_s10), na.rm = TRUE),
    
    # do more stages sparation like 3 stages: early, mid and late.
    # here is gdd
    gdd_early  = sum(c_across(gdd_s1:gdd_s3), na.rm = TRUE),
    gdd_mid    = sum(c_across(gdd_s4:gdd_s7), na.rm = TRUE),
    gdd_late   = sum(c_across(gdd_s8:gdd_s10), na.rm = TRUE),
    # do prcp also
    prcp_early = sum(c_across(prcp_s1:prcp_s3), na.rm = TRUE),
    prcp_mid   = sum(c_across(prcp_s4:prcp_s7), na.rm = TRUE),
    prcp_late  = sum(c_across(prcp_s8:prcp_s10), na.rm = TRUE)
  ) %>%
  ungroup()

```

```{r save weather stage test}
write_csv(weather_test_stage_wide2,
          "../data/weather_test_stage_wide.csv")
```

## f. Add growth period to test data
```{r add test growth period}

# the weather wide range has been filter by growth of train, so we have to use raw 2024 year data for test data
test_mer <- read_csv("../data/testing/testing_merged.csv")
train_mer <- read_csv("../data/training/training_merged.csv")
weather_raw<- read_csv("../data/weather_allsites_12years.csv") %>%
  clean_names()

# use train data to reverse create growth period for test data
# the plant and harvest time mostly depend on site, rather than hybrid.
site_stage_info <- train_mer %>%
  mutate(
    date_planted   = mdy(date_planted),
    date_harvested = mdy(date_harvested),
    plant_doy      = yday(date_planted),
    harvest_doy    = yday(date_harvested)
  ) %>%
  group_by(site) %>%
  summarise(
    plant_doy_mean   = round(mean(plant_doy, na.rm = TRUE)),
    harvest_doy_mean = round(mean(harvest_doy, na.rm = TRUE)),
    .groups = "drop"
  )

# add plant and harvest to test 
test_stage_base <- test_mer %>%
  left_join(site_stage_info, by = "site") %>%
  mutate(
    date_planted = as.Date(plant_doy_mean - 1,
                           origin = paste0(year, "-01-01")),
    date_harvested = as.Date(harvest_doy_mean - 1,
                             origin = paste0(year, "-01-01"))
  )

# check the site didn't added by missing
test_stage_base %>%
  filter(is.na(plant_doy_mean) | is.na(harvest_doy_mean)) %>%
  distinct(site)
# ONH3 got missing.

test_stage_base %>%
  count(site) %>%
  mutate(prop = n / sum(n)) %>%
  arrange(desc(prop))
# ONH3 take up 4.2% of whole  data. So we can't ignore it.

# ONH3 never shows in train files.
site_ONH3 <- test_stage_base %>%
  filter(site == "ONH3") %>%
  distinct(site,latitude,longitude)
# 45.38412,75.7129

```
```{r prepare for download }
site_coord_test <- test_stage_base %>%
  group_by(site) %>%
  summarise(
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    .groups = "drop"
  ) %>%  
  # remove na lat and lon
  filter(!is.na(latitude), !is.na(longitude))
```

```{r download test weather data}
weather_test<- pmap_dfr(
  list(site_coord_test$longitude, site_coord_test$latitude, site_coord_test$site),
  get_weather
)
  #beep
  beep()
  
  # save data 
  write_csv(weather_test, "../data/weather_test.csv")
```
Then we need to do the same steps to test data as what we do to train data.
But there's one problem, we assume the plant and growth based on site info, but ONH3 never shows in train data set. Notice:
DEH1 → Delaware
GAH1 → Georgia
IAH1 → Iowa
INH1 → Indiana
MNH1 → Minnesota
MOH1 → Missouri
NCH1 → North Carolina
NEH1 → Nebraska
NYH1 → New York
ONH1 → Ohio
So we can only assume ONH3 in ohio, and same plant time in same state.And there OHH1 shows both in train and test set. Let's make this to ONH3

```{r populate the growth date for test}

site_stage_info <- train_mer %>%
  mutate(
    date_planted = mdy(date_planted),
    date_harvested = mdy(date_harvested),
    plant_doy = yday(date_planted),
    harvest_doy = yday(date_harvested)
  ) %>%
  group_by(site) %>%
  summarise(
    plant_doy_mean = round(mean(plant_doy, na.rm = TRUE)),
    harvest_doy_mean = round(mean(harvest_doy, na.rm = TRUE)),
    .groups = "drop"
  )

oh_stage_info <- train_mer %>%
  filter(site == "OHH1") %>%
  mutate(
    date_planted = mdy(date_planted),
    date_harvested = mdy(date_harvested),
    plant_doy = yday(date_planted),
    harvest_doy = yday(date_harvested)
  ) %>%
  summarise(
    plant_doy_oh = round(mean(plant_doy, na.rm = TRUE)),
    harvest_doy_oh = round(mean(harvest_doy, na.rm = TRUE))
  )

test_stage_base <- test_mer %>%
  left_join(site_stage_info, by = "site") %>%
  mutate(
    plant_doy_final = case_when(
      site == "ONH3" & is.na(plant_doy_mean) ~ oh_stage_info$plant_doy_oh,
      TRUE ~ plant_doy_mean
    ),
    harvest_doy_final = case_when(
      site == "ONH3" & is.na(harvest_doy_mean) ~ oh_stage_info$harvest_doy_oh,
      TRUE ~ harvest_doy_mean
    ),
    date_planted = as.Date(plant_doy_final - 1, origin = paste0(year, "-01-01")),
    date_harvested = as.Date(harvest_doy_final - 1, origin = paste0(year, "-01-01"))
  )
test_stage_base %>%
  filter(site == "ONH3") %>%
  select(site, year, date_planted, date_harvested) %>%
  distinct()

 # save data 
  write_csv(test_stage_base, "../data/testing/test_stage_base.csv")
```

# 5) Final check before train model
We can start train model at this stage, but we may check the data so that we can decide the further direction. or whether we need to do more data wrangling about the current dataset.

## a. Check all 3 csv files
```{r check 3 csv files}
train_mer <- read_csv("../data/training/training_merged.csv")
# 164k rows, 17 cols
test_mer <- read_csv("../data/testing/test_stage_base.csv")
# 10k rows, 17 cols

weather_train_fe <- read_csv("../data/weather_stage_wide.csv")
# 262 rows, 74 cols
weather_test_fe <- read_csv("../data/weather_test_stage_wide.csv")
# 23 rows, 74 cols

# Let's see the difference in two merged data.
setdiff(names(train_mer), names(test_mer))
# so we don't have these cols in test data
# [1] "detailed_site"  "replicate"      "block"          "grain_moisture"

setdiff(names(test_mer), names(train_mer))
# [1] "plant_doy_mean"    "harvest_doy_mean"  "plant_doy_final"  
# [4] "harvest_doy_final"
 
```
```{r select cols for train and test data}
train_mer2 <- train_mer %>%
  dplyr::select(-detailed_site, 
                -replicate, 
                -block,
                -date_planted,
                -date_harvested,
                -grain_moisture,
                -latitude,
                -longitude,)
# we do the GDD and stage stuff, so let's remove plant and harvest date.
# rep and block not in the test, only depend on site name, remove.
# what about moisture, it's also predicted values of corn grain, not in the soil. remove.
# what about lon and lat? We have the weather according to site, so remove.

# let's keep same cols for test merged data
test_mer2 <- test_mer %>%
  dplyr::select(-plant_doy_mean,
                -harvest_doy_mean,
                -plant_doy_final,
                -harvest_doy_final,
                -latitude,
                -longitude,
                -date_harvested,
                -date_planted)

setdiff(names(train_mer2), names(test_mer)) 
# no col difference now. Finally!
```

```{r move weather to merged data}
train_final <- train_mer2 %>%
  left_join(weather_train_fe, by = c("site", "year"))
# no more rows, 81 cols, means good!

test_final <- test_mer2 %>%
  left_join(weather_test_fe, by = c("site", "year"))
# no more rows, 81 cols, means good!

```



## b. Export files.
```{r Export files}
# let's save 
write_csv(train_final,
          "../data/training/train_final.csv")
saveRDS(train_final, "../data/training/train_final.rds")


# to read rds file.
train_final2 <- readRDS("../data/training/train_final.rds")
# checked, it's compressed into rds file to reduce the size of train file.


# save test data
write_csv(test_final,
          "../data/testing/test_final.csv")

```

So,
train_final.rds
test_final.csv 
will be our last 2 files.