---
title: "Data Wrangling"
format: html
editor: Ming/Ishwari
---

# 1) Setup

Prepare the packages and raw materials that we need for data wrangling.

```{r Setup, message=FALSE}
# Loading packages (every time you open R)
#install.packages("nasapower") #
#install.packages("beepr")
library(readxl) # to read excel files
library(janitor) # to clean data
library(dplyr) # wranglingdd
library(tidyr) # wrangling
library(readr) # to export csv
library(nasapower) # for NASA POWER API
library(purrr) # for mapping
library(beepr) # for beep
library(stringr)
.libPaths()
```

```{r Read files}
getwd() #get your working directory
Meta <- read_csv("../data/training/training_meta.csv")
Soil <- read_csv("../data/training/training_soil.csv")
Trait <- read_csv("../data/training/training_trait.csv")

glimpse(Meta)
glimpse(Soil)
glimpse(Trait)

```


# 2) Open source data


```{r test api }
# get weather data from NASA Power based on year, site, lon, lat in meta data

#meta_unique <- Meta %>%
#  distinct(site, year, latitude, longitude)


site_coord <- meta_cleaned %>%
  group_by(site) %>%
  summarise(
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    .groups = "drop"
  )
glimpse(meta_unique) 
# Example of R api 
#https://docs.ropensci.org/nasapower/articles/nasapower.htm

# learn api
help(package = "nasapower")



# test api
weather_test <- get_power(
  community = "AG",   # agriculture
  lonlat = c(-83.4, 35.2),  # 经度, 纬度
  dates = c("2014-01-01", "2023-12-31"),
  pars = c(
  "T2M",
  "T2M_MAX",
  "T2M_MIN",
  "PRECTOTCORR",
  "ALLSKY_SFC_SW_DWN"
)  # mean tmp, max, min, precipitation, radiation
)

glimpse(weather) 
```
```{r create a function}
get_weather <- function(lon, lat, yr, site_name ) {
  cat("Running:", site_name, yr, "\n")
  x <- get_power(
    community = "AG",
    lonlat = c(lon, lat),
    dates = c(paste0(yr, "-01-01"), paste0(yr, "-12-31")),
    pars = c(
      "T2M",
      "T2M_MAX",
      "T2M_MIN",
      "PRECTOTCORR",
      "ALLSKY_SFC_SW_DWN"
    ),
    temporal_api = "daily"
  )
  
  Sys.sleep(1)  # avoid be banned by api limit 
   return(x) # return weather data frame
}


```

```{r test func}
weather_test_func <- meta_unique %>% slice(1:5) %>%
  mutate(
    weather = pmap(
      list(longitude, latitude, year, site),
      ~ get_weather(..1, ..2, ..3, ..4)
    )
  ) %>%
  tidyr::unnest(weather)

```

```{r iteration}
weather_all_year <- meta_unique %>%
  mutate(
    weather = pmap(
      list(longitude, latitude, year, site),
      ~ get_weather(..1, ..2, ..3, ..4)
    )
  ) %>%
  tidyr::unnest(weather)
 #beep
  beep()
```
```{r check the number of site for each year}
meta_unique %>%
  count(year)
```

```{r save as csv}
write_csv(weather_all_year, "../data/weather_raw.csv")
```

# 3) Data wrangling  

## On training data
```{r clean_names }
meta_cleaned <- clean_names(Meta)
soil_cleaned <- clean_names(Soil)
trait_cleaned <- clean_names(Trait)

glimpse(soil_cleaned)

```

Trying to merge three csv. But lots of problems on site-year. So we have to deal with it before merge.
Different code plus num(e.g TXH1, TXH2), means they are different site, means different lan and lon.

later suffix either lower case a/b or late/early(e.g "MOH1a", "MOH1-rep2") means different rep/block info, but doesn't affect the lan and lon info.

```{r handling site-year}
# clean soil, separate site-year
soil_separate <- soil_cleaned %>%
  # "_(?=[^_]+$)" means look for the last "_" from right side to do the separation.
  separate(site, into = c("site", "year"), sep = "_(?=[^_]+$)") %>%
  mutate(year = as.integer(year))


# trait,soil, and meta all have different site-year. check the number
trait_key <- trait %>%
  distinct(site, year)
# 270 rows

meta_key <- meta %>%
  distinct(site, year)
# 251 rows

soil_key <- soil_separate %>%
  distinct(site, year)
# 186 keys


# avoid many-to-many merge
## check meta lat,lon base on site-year info
meta_cleaned %>%
  count(site, year) %>%
  filter(n > 1)

meta_check_site_year<-meta_cleaned %>%
  group_by(site, year) %>%
  summarise(n = n(),
            n_distinct_lat = n_distinct(latitude),
            n_distinct_lon = n_distinct(longitude))
## lan and lon all same, so we can use distinct as long as they're are from same site code, even if the lan and lon differ slightly. But we have to choose one set of lan and lon from the same site code, to avoid the many-to-many merge problem.

# average the lon and lat for the same site code
meta_unique <- meta_cleaned %>%
  group_by(site, year) %>%
  summarise(
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    .groups = "drop"
  )

# do the same to soil data
soil_unique <- soil_separate %>%
  group_by(site, year) %>%
  summarise(
    soilp_h = mean(soilp_h, na.rm = TRUE),
    om_pct = mean(om_pct, na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups = "drop"
  )

# now we have the soil and meta data with distinct site year so that we can avoid that many-to-many problem when merge to trait data 

# set same site for 3 csv

# test regular expression
sub("^([A-Za-z]+[0-9]+).*", "\\1",
    c("TXH1-Late", "MOH1a", "MOH1-rep2", "IAH1"))
#  "TXH1" "MOH1" "MOH1" "IAH1"

trait2 <- trait %>%
  rename(detailed_site = site) %>%
  mutate(
    site = detailed_site %>%
    sub("^([A-Za-z]+[0-9]+).*", "\\1", x = .) # remove a/b/c,rep, later/early
      
  ) %>% 
  # move to 3rd col
  relocate(site, .after = detailed_site)

# check trait2, row #943, IAH1a -> IAH1, it worked.
# row # 49901 2017 TXH1-Early -> TXH1, it worked.

```
Now, we make the site clearly separate from detailed site in trait data.
And we make site unique in meta and soil data.
We can start merge.

```{r  merge data & check }

# merge
data_merged <- trait2 %>%
  left_join(meta_unique, by = c("site" = "site", "year" = "year")) %>%
  left_join(soil_unique, by = c("site" = "site", "year" = "year"))

# check many-to-many problem
nrow(trait2) == nrow(data_merged)
# True means we did it right, there's no many-to-many problem happened.


# overwrite with rep info in soil data
soil_rep <- soil_separate %>%
  mutate(
    site_base = str_remove(site, "_\\d+$"),
    rep_soil = as.integer(str_extract(site, "(?<=_)\\d+$"))
  ) %>%
  filter(!is.na(rep_soil)) %>%
  select(site_base, year, rep_soil, soilp_h, om_pct, soilk_ppm, soilp_ppm)

data_fix <- data_merged %>%
  left_join(
    soil_rep,
    by = c("site" = "site_base", "year" = "year", "replicate" = "rep_soil"),
    suffix = c("", "_rep")
  )

data_fix <- data_fix %>%
  mutate(
    soilp_h = coalesce(soilp_h_rep, soilp_h),
    om_pct = coalesce(om_pct_rep, om_pct),
    soilk_ppm = coalesce(soilk_ppm_rep, soilk_ppm),
    soilp_ppm = coalesce(soilp_ppm_rep, soilp_ppm)
  ) %>%
  select(-ends_with("_rep"))
# read the row#63403 &63404, the soil is different now. Means we give the different rep to trait data.

# save as merged data
write_csv(data_fix, "../data/training/training_merged.csv")
```
This means we merge correctly, even though we lack some site-year info initially from meta and soil data. 

Then we can do some data wrangling & feature engineering on weather data.
## On testing data
We should do the same to testing data as to training data.
```{r clean_names }
test_soil_cleaned <- read_csv("../data/testing/testing_soil.csv") %>%
  clean_names()
test_meta_cleaned <- read_csv("../data/testing/testing_meta.csv") %>%
  clean_names()
test_submission_cleaned <- read_csv("../data/testing/testing_submission.csv") %>%
  clean_names()



glimpse(test_soil_cleaned)

```

```{r separate year and site for soil and meta}
test_soil_sep <- test_soil_cleaned %>%
  # "_(?=[^_]+$)" means look for the last "_" from right side to do the separation.
  separate(site, into = c("site", "year"), sep = "_(?=[^_]+$)") %>%
  mutate(year = as.integer(year)) 

# get unique site year, by average
test_soil_sep2 <- test_soil_sep %>%
  group_by(site, year) %>%
  summarise(
    soilp_h = mean(soilp_h, na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    om_pct = mean(om_pct, na.rm = TRUE),
    .groups = "drop"
  )
# now it's 16 rows


# test_meta all rows unique, so we don't need this step for test_meta


```

```{r merge & check}
# merge
test_data_merged <- test_submission_cleaned %>%
  left_join(test_soil_sep2, by = c("site" = "site", "year" = "year")) %>%
  left_join(test_meta_cleaned, by = c("site" = "site", "year" = "year"))
# check many-to-many problem
nrow(test_data_merged) == nrow(test_submission_cleaned)
# True means we did it right, there's no many-to-many problem happened.

 
# save as merged data
write_csv(test_data_merged, "../data/testing/testing_merged.csv")
```

# 4) Feature engineering on weather data
## Download again
Based on updated view on the merged trait data, we have to download weather again for all the unique site-year across whole 2013-2024 12 years.
```{r read the data,check the site-year match }
#check the distint site-year of weather
weather <- read.csv("../data/weather_raw.csv")
weather_unique_siteyear <- weather %>%
  distinct(site, year) 
# 251 rows

#check the distint site-year of merged data
data_merged <- read.csv("../data/training/training_merged.csv")
train_data_unique_siteyear<-  data_merged %>%
  distinct(site, year)
# 264 rows

# then we need to download again weather data about all the unique site 
train_data_unique_site <- data_merged %>%
  distinct(site, latitude,longitude)
# 39 rows,
# we need all these site across 2014-2023 years 
site_coord <- data_merged %>%
  group_by(site) %>%
  summarise(
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    .groups = "drop"
  ) %>%  
  # remove na lat and lon
  filter(!is.na(latitude), !is.na(longitude))


```


```{r create a function}
get_weather <- function(lon, lat, site_name) {
  cat("Running:", site_name, "\n")
  
  x <- get_power(
    community = "AG",
    lonlat = c(lon, lat),
    dates = c("2013-01-01", "2024-12-31"),
    pars = c(
      "T2M",
      "T2M_MAX",
      "T2M_MIN",
      "PRECTOTCORR",
      "ALLSKY_SFC_SW_DWN"
    ),
    temporal_api = "daily"
  )
  
  x$site <- site_name
  
  Sys.sleep(0.5)
  
  return(x)
}


```

```{r test func}
weather_test_func <- site_coord %>%
  slice(1:5) %>%
  mutate(
    weather = pmap(
      list(longitude, latitude, site),
      ~ get_weather(..1, ..2, ..3)
    )
  ) %>%
  tidyr::unnest(weather)
  #beep
  beep()
```

```{r iteration}
weather_raw <- pmap_dfr(
  list(site_coord$longitude, site_coord$latitude, site_coord$site),
  get_weather
)
  #beep
  beep()
  
  # save data 
  write_csv(weather_raw, "../data/weather_allsites_12years.csv")
```
Until now, we get the new weather data, let's start feature engineering. 
## Feature enineering
```{r EDA for the training merged data }


```

```{r accumulative temp }


```