---
title: "XG-Boost_Training"
format: html
editor: Ming/Ishwari
---

# 1) Setup

```{r}
# shared_lib_path <- "/work/crss8030/instructor_data/shared_R_libs"

#class_packages <- c("tidymodels", "tidyverse", "vip", "ranger", "finetune", "parsnip", "reticulate", "xgboost", "doParallel", "lme4", "here")

#install.packages(class_packages, lib = shared_lib_path)
```

# Locate the installed libraries (important step while running the script in cluster - sapelo2)

```{r}
.libPaths(c("/work/crss8030/instructor_data/shared_R_libs", .libPaths()))
```
```{r}
#.libPaths()
```


```{r}
#| message: false
#| warning: false
library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(ranger)       # Fast implementation of Random Forests
library(tidyverse)    # Data wrangling and visualization
library(tidymodels)   # for func：initial_split
library(doParallel)   # For parallel computing (useful during resampling/tuning)
library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
library(lme4)         # The matrix package and other mathmatical computation for xgboost
library(here)         # Gives the absolute path starting from Rproj, works well with different scripts (R, .rmd or .qmd)
library(quarto)       # helps to export the qmd script to R script
#library(caret)       # Other great library for Machine Learning 

here::here()
```


```{r read train data}
corn <- readRDS("../data/training/train_final.rds")
head(corn)
```

# 2) ML workflow

## 1. Pre-processing

### a. Data split
```{r corn_split}
set.seed(123) # Setting seed to get reproducible results 
corn_split <- initial_split(
  corn, 
  # proportion of split, more for test data set cuz we have lots of rows
  prop = .65, 
  strata = yield_mg_ha  # Stratify by target variable
  )
corn_split
# <Training/Testing/Total>
# <106908/57569/164477>
```


```{r corn_train}
corn_train <- training(corn_split)  # 65% of data
corn_train #This is your training data frame

corn_test <- testing(corn_split)    # 35% of data
corn_test

```


### b. Distribution of target variable

```{r distribution}
density_plot <- ggplot() +
  geom_density(data = corn_train, 
               aes(x = yield_mg_ha),
               color = "red") +
  geom_density(data = corn_test, 
               aes(x = yield_mg_ha),
               color = "blue") 
  # almost perfect distribution

# lets print the undergoing process in the console, (important when running in the terminal later on)
cat(paste0("Saving...... density plot of train and test data"))

ggsave(plot = density_plot, 
       path = here("output", "png"),
       filename = "density_plot_test_train.png",
       height = 6,
       width = 9,
       dpi = 600)
```

## 2. Data processing with recipe
```{r corn_recipe}
cat("\nCreate recipe for data preprocessing\n")
# Create recipe for data preprocessing
corn_recipe <- recipe(yield_mg_ha ~ ., data = corn_train) %>% # Remove identifier columns and months 
  step_rm(
    year,       # Remove year identifier
    site,        # Remove site identifier
    hybrid,      #Remove hybrid, cuz too many cols 
    parent1      #parent1
  ) %>%
  step_string2factor(parent2) %>%
  step_unknown(parent2) %>%
  step_unknown(previous_crop) %>%
  step_other(parent2, threshold = 0.01, other = "rare")%>%
  step_dummy(all_nominal_predictors())  # make cat to num for hybrid and pervious crop
corn_recipe
```

```{r corn_prep}
# Prep the recipe to estimate any required statistics
corn_prep <- corn_recipe %>% 
  prep()

# Examine preprocessing steps
corn_prep
```

## 3. Training

### a. Model specification

```{r xgb_spec}
xgb_spec <- #Specifying XgBoost as our model type, asking to tune the hyperparameters
  boost_tree(
   # Total number of boosting iterations
    trees = tune(),
         # Maximum depth of each tree
    tree_depth = tune(),
             # Minimum samples required to split a node
    min_n = tune(),
        # Step size shrinkage for each boosting step
    learn_rate = tune()
    ) %>%
        #specify engine 
  set_engine("xgboost") %>%
       # Set to mode
  set_mode("regression")
xgb_spec

```


### b. Cross-validation setup
We use 10-fold cross-validation to evaluate model performance during tuning:
```{r}
set.seed(1234)  
resampling_foldcv <- vfold_cv(corn_train, # Create 10-fold cross-validation resampling object from training data
                              v = 10)


# Create leave one year out cv object from the sampling data
resampling_fold_loyo<-group_vfold_cv(corn_train,
                                     group = year
                                     )

# Create leave one location out cv object from the sampling data
resampling_fold_loso<-group_vfold_cv(corn_train,
                                     group = site
                                     )
```
### c. Hyperparameter grid with Latin Hypercube Sampling

We use Latin hypercube sampling to generate a diverse grid of hyperparameter combinations:

```{r }
xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  learn_rate(),
  trees(),
  # 10 for test on our desktop, 1000 for HPC
  size = 1
)

xgb_grid
```
See the combinations
```{r}
ggplot(data = xgb_grid,
       aes(x = tree_depth, 
           y = min_n)) +
  geom_point(aes(color = factor(learn_rate),
                 size = trees),
             alpha = .5,
             show.legend = FALSE)
```

## 4. Model Tuning
Detect core numbers of Sapelo2
```{r}
# Detect Cores for Slurm (Cluster - Sapelo2)

# lets look into our current environment to see the available cores
n_cores <- as.numeric(Sys.getenv("SLURM_CPUS_ON_NODE"))

if (is.na(n_cores)) n_cores <- parallel::detectCores() - 1

# Start the Cluster
cl <- makePSOCKcluster(n_cores)

registerDoParallel(cl)

cat(paste0("\nFound and registered ", n_cores, " cores to work with\n"))

```


```{r xgb_grid_result}

# Create the list of CV techniques so we can loop through them
cv_list <- list(
  vfold = resampling_foldcv,
  year = resampling_fold_loyo,
  site = resampling_fold_loso
  
)

# Create a emty list to store the results from the loop
results <- list()


# Create our loop
for (i in seq_along(cv_list)) {
  name <- names(cv_list)[i]
  
  results[[name]] <- tune_race_anova(object = xgb_spec, 
                                     preprocessor = corn_recipe,
                                     resamples = cv_list[[i]],
                                     grid = xgb_grid,
                                     control = control_race(save_pred=TRUE,
                                                             parallel_over = "everything"
                                                             )
                                     )
}

beepr::beep()
stopCluster(cl)
results

```


Let's check the results
```{r}
library(dplyr)

# Create a dataframe structure from the list obtained after running the loop
results_df <-  tibble(method = names(results),
                      diff_cv = results
                      ) %>%
  # Collect the metrices for each CV techniques using map function
  mutate(metrices = map2(diff_cv,method,
                         ~.x %>%
                           collect_metrics() %>%
                           mutate(method = .y, .before= "trees")
                           
                         ))

results_df 

# bind all the metrices together so to select the best performing one
all_metrices <- do.call(bind_rows, results_df$metrices)
```

```{r}
# Automating to pull the best method out of 3 we ran

best_method <- all_metrices %>%
  filter(.metric == "rmse") %>%
  slice_min(mean, n = 1) %>%
  pull(method)
  

# Getting the metrice (hyperparameter values of the best performing CV)
best_cv_object <- results_df %>%
  filter(method == best_method) %>%
  pull(diff_cv) %>% 
  first()
  
```

## 5. Select Best Models

We select the best models using three strategies (best, within 1 SE, within 2% loss) which we learned in class:

```{r}
# Best RMSE
best_rmse <- best_cv_object %>% 
      select_best(metric = "rmse")%>% 
  mutate(source = "best_rmse")

best_rmse
```

```{r}
# Based on greatest R2
best_r2 <- best_cv_object %>% 
  select_best(metric = "rsq")%>% 
  mutate(source = "best_r2")

best_r2
```

## 6. Compare and Finalize Model

```{r comparing values}
best_rmse %>% 
  bind_rows(best_rmse, 
            best_r2) %>%
  dplyr::select(source, everything())
```
  
Record the best combo here: 
trees = 116
tree_depth = 9
min_n = 34
learn_rate = 0.01690976

## 7. Final Specification

```{r final_spec_fit}
final_spec <- boost_tree(
  trees = best_r2$trees,           # Number of boosting rounds (trees)
  tree_depth = best_r2$tree_depth, # Maximum depth of each tree
  min_n = best_r2$min_n,           # Minimum number of samples to split a node
  learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

final_spec
```

## 8. Final Fit and Predictions

### Validation

```{r final_fit}
set.seed(10)
final_fit <- last_fit(final_spec,
                weather_recipe,
                split = weather_split)

final_fit %>%
  collect_predictions()
```


## 9. Evaluate on Test Set

```{r final_fit_metrics}
test_met <- final_fit %>%
  collect_metrics() %>% 
  mutate(estimate = round(.estimate, 3))
```

## 10. Evaluate on Training Set

```{r}
final_spec %>%
  fit(strength_gtex ~ .,
      data = bake(weather_prep, 
                  weather_train)) %>%
  augment(new_data = bake(weather_prep, 
                          weather_train)) %>% 
  rmse(strength_gtex, .pred) %>%
  bind_rows(
    
    
# R2
final_spec %>%
  fit(strength_gtex ~ .,
      data = bake(weather_prep, 
                  weather_train)) %>%
  augment(new_data = bake(weather_prep, 
                          weather_train)) %>% 
  rsq(strength_gtex, .pred))
```

```{r}
final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = strength_gtex,
             y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  scale_x_continuous(limits = c(1, 25)) +
  scale_y_continuous(limits = c(1, 25)) 
```

## 11. Predicted vs Observed Plot (publication ready)

```{r}
publication_ready <- final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = strength_gtex,
             y = .pred)) +
  # Changing the shape of the points to 
  geom_point(aes(fill = strength_gtex),shape = 21, alpha = .7, show.legend = F) +
  # Using colorblind friendly color scale 
  scale_fill_viridis_c(option = "H" ) +
  # Changing the color and linetype of abline
  geom_abline(color = "red", linetype = 2) +
  geom_smooth(method = "lm") +
  labs(x = "Observed Strenght (gtex)",
       y = "Predicted Strength (gtex)")+
  # Annotate R-sq and RMSE value in the plot
  annotate("label", x = Inf, y = -Inf,
           label = paste0("R-sq:",test_met$estimate[[2]], "\nRMSE:",test_met$estimate[[1]]),
          hjust = 1, vjust = -0.5 ) +
  
  
  scale_x_continuous(limits = c(20, 40)) +
  scale_y_continuous(limits = c(20, 40)) +
  
  # Changing the theme of the plot
  theme(panel.background = element_rect(fill = "gray82"),
        panel.grid = element_blank()
        )

publication_ready

cat(paste0("\nSaving...... publication ready plot for test dataset\n"))

ggsave(plot = publication_ready, 
       path = here("output", "png"),
       filename = "model_perf_test_data.png",
       height = 6,
       width = 9,
       dpi = 600)
  
```


## 12. Variable Importance

```{r final_spec}
vip <- final_spec %>%
  fit(strength_gtex ~ .,
         data = bake(weather_prep, weather_train)) %>% #There little change in variable improtance if you use full dataset
    vi() %>%
  mutate(
    Variable = fct_reorder(Variable, 
                           Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable,
             fill = Importance)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)+
  theme(panel.background  = element_rect(fill = "gray82"),
        panel.grid = element_blank())

cat(paste0("\nSaving...... publication ready plot for variable importance\n"))

ggsave(plot = vip, 
       path = here("output/png"),
       filename = "vip_test_data.png",
       height = 6,
       width = 9,
       dpi = 600)
```

```{r covert qmd to R script}
knitr::purl("XG-Boost_Training.qmd", output = "sapelo_script.R", documentation = 0)
```